---
title: "Covid 19 Death Prediction Model 20200315"
author: "Josh Harwood"
date: "March 15, 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


For this problem, I modeling the effect on the log of daily new confirmed cases to the log of daily new deaths using transfer function analysis (autoregression).

```{r message=FALSE, warning=FALSE, results='hide'}
library(readr)

# Confirmed
download.file("https://data.humdata.org/hxlproxy/api/data-preview.csv?url=https%3A%2F%2Fraw.githubusercontent.com%2FCSSEGISandData%2FCOVID-19%2Fmaster%2Fcsse_covid_19_data%2Fcsse_covid_19_time_series%2Ftime_series_covid19_confirmed_global.csv&filename=time_series_covid19_confirmed_global.csv","confirmed.csv")

# deaths
download.file("https://data.humdata.org/hxlproxy/api/data-preview.csv?url=https%3A%2F%2Fraw.githubusercontent.com%2FCSSEGISandData%2FCOVID-19%2Fmaster%2Fcsse_covid_19_data%2Fcsse_covid_19_time_series%2Ftime_series_covid19_deaths_global.csv&filename=time_series_covid19_deaths_global.csv","deaths.csv")

# recovered
download.file("https://data.humdata.org/hxlproxy/api/data-preview.csv?url=https%3A%2F%2Fraw.githubusercontent.com%2FCSSEGISa
ndData%2FCOVID-19%2Fmaster%2Fcsse_covid_19_data%2Fcsse_covid_19_time_series%2Ftime_series_covid19_recovered_global.csv&filen
ame=time_series_covid19_recovered_global.csv"
,"recovered.csv")

df.confirmed = read_csv("confirmed.csv")
df.deaths = read_csv("deaths.csv")
df.recovered = read_csv("recovered.csv")
```

```{r}
library(tidyverse)
# US
# confirmed
df.confirmed.grp <- df.confirmed %>%
  mutate(country= as.factor(`Country/Region`)) %>%
  filter(country == "US", ) %>%
  gather(key="date", value="confirmed", 5:94) %>%
  select(-`Province/State`,-`Country/Region`,-`Lat`,-`Long`) %>%
  mutate(date=as.Date(date,"%m/%d/%y")) %>%
  group_by(country, date) %>%
  arrange(date) %>%
  summarize(confirmed = sum(confirmed)) %>%
  mutate(daily.confirmed = confirmed-dplyr::lag(confirmed,n=1)) %>%
  mutate(log.daily.confirmed = log(1+confirmed-dplyr::lag(confirmed,n=1),base= exp(1))) %>%
  select(date,log.daily.confirmed) %>%
  filter(date>=as.Date("03/01/2020","%m/%d/%y")) #This is when cases trended in a consistent manner

# deaths
df.deaths.grp <- df.deaths%>%
  mutate(country= as.factor(`Country/Region`)) %>%
  filter(country == "US") %>%
  gather(key="date", value="deaths", 5:94) %>%
  select(-`Province/State`,-`Country/Region`,-`Lat`,-`Long`) %>%
  mutate(date=as.Date(date,"%m/%d/%y")) %>%
  group_by(country, date) %>%
  arrange(date) %>%
  summarize(deaths = sum(deaths)) %>%
  mutate(daily.deaths = deaths-dplyr::lag(deaths,n=1)) %>%
  mutate(log.daily.deaths = log(1+deaths-dplyr::lag(deaths,n=1),base= exp(1))) %>%
  select(date,log.daily.deaths) %>%
  filter(date>=as.Date("03/01/2020","%m/%d/%y")) #This is when cases trended in a consistent manner

# recovered
df.recovered.grp <- df.recovered%>%
  mutate(country= as.factor(`Country/Region`)) %>%
  filter(country == "US") %>%
  gather(key="date", value="recovered", 5:94) %>%
  select(-`Province/State`,-`Country/Region`,-`Lat`,-`Long`) %>%
  mutate(date=as.Date(date,"%m/%d/%y")) %>%
  group_by(country, date) %>%
  arrange(date) %>%
  summarize(recovered = sum(recovered)) %>%
  mutate(daily.recovered = recovered-dplyr::lag(recovered,n=1)) %>%
  mutate(log.daily.recovered = log(1+recovered-dplyr::lag(recovered,n=1),base= exp(1))) %>%
  select(date,log.daily.recovered) %>%
  filter(date>=as.Date("03/01/2020","%m/%d/%y")) #This is when cases trended in a consistent manner
```

Compare Graphs
```{r fig.width=9}
plot(df.confirmed.grp$date, df.confirmed.grp$log.daily.confirmed,type='l',lty=1,col='black',
     ylim=c(0,10))
lines(df.deaths.grp$date, df.deaths.grp$log.daily.deaths, lty = 1, col='red')
lines(df.recovered.grp$date, df.recovered.grp$log.daily.recovered, lty = 1, col='blue')
```

Generate a time series. Certainly none of these series is stationary currently.
```{r}
library(TSA)
library(lubridate)
# Confirmed
confirmed.ts.prep = df.confirmed.grp[,3]
covid.confirmed.ts = ts(data=confirmed.ts.prep, start = decimal_date(as.Date("2020-03-15")), 
                    frequency = 365.25)
#365.25 accounts for leap year
plot(covid.confirmed.ts)
# decimal_date() allows me to start on a specific date

# Determine Prewhitening formula for confirmed Cases
acf(as.vector(covid.confirmed.ts))
pacf(as.vector(covid.confirmed.ts))
# Appears to be AR(1) model
mod1.conf <- arima(covid.confirmed.ts, order = c(3,1,3), 
                 seasonal = list(order=c(0,0,0))) 
mod1.conf
plot(window(rstandard(mod1.conf), start = decimal_date(as.Date("2020-03-15"))), 
     ylab = 'Standardized Residuals', type = 'o')
abline(h=0)
# ACF of Standardized Residuals
acf(as.vector(rstandard(mod1.conf)), lag.max = 40,ci.type="ma")
pacf(as.vector(rstandard(mod1.conf)), lag.max = 40)
# qqplot
qqnorm(window(rstandard(mod1.conf), start = decimal_date(as.Date("2020-03-15"))))
qqline(window(rstandard(mod1.conf), start = decimal_date(as.Date("2020-03-15"))))
Box.test(residuals(mod1.conf),lag=30,type=c("Ljung-Box"))
```
Looking at the first 30 lags our model residuals fail to reject the null hypothesis of the Ljung-Box test the the residuals are independently distributed with no serial autocorrelation.


```{r}
# Transfer Function for Deaths
# Deaths
deaths.ts.prep = df.deaths.grp[,3]
covid.deaths.ts = ts(data=deaths.ts.prep, start = decimal_date(as.Date("2020-03-15")), 
                    frequency = 365.25) 
conf.death.ts = ts.intersect(covid.confirmed.ts, covid.deaths.ts)

plot(conf.death.ts,yax.flip=T)

library(forecast)
fitwhite.x <- residuals(mod1.conf)
fitwhite.y <- residuals(Arima(covid.deaths.ts, model=mod1.conf))
print(ccf(as.vector(fitwhite.x),as.vector(fitwhite.y), ylab='ccf'))

```

We see significant positive cross-correlation after pre-whitening which indicates that the two series are likely not spuriously correlated. 

```{r}
conf.data <- df.confirmed.grp %>%
  select(date, log.daily.confirmed)
death.data <- df.deaths.grp %>%
  select(date, log.daily.deaths)
death.mod.data <- conf.data %>%
  left_join(death.data, by= 'date') %>%
  select(date,log.daily.confirmed, log.daily.deaths) %>%
  mutate(lag14.log.confirmed = lag(log.daily.confirmed, n=14),
         lag7.log.confirmed = lag(log.daily.confirmed, n=7)) %>%
  filter(date >= as.Date("03/15/2020","%m/%d/%y"))

# Only lag 7 was significant
death.mod1=lm(log.daily.deaths~lag14.log.confirmed+lag7.log.confirmed,
              data=death.mod.data)
summary(death.mod1)
acf(residuals(death.mod1),ci.type='ma')
pacf(residuals(death.mod1))
# eacf(residuals(death.mod1))
qqnorm(window(rstandard(death.mod1)))
qqline(window(rstandard(death.mod1)))
Box.test(residuals(death.mod1),lag=30,type=c("Ljung-Box"))


# Don't need this, just lagged regression noq
# # Try ARMA(3,3) for Zt
# # Turned out that all terms are not significant except the ar2
# death.autoreg <- arima(death.mod.data$log.daily.deaths, order=c(1,0,1),
#                        xreg=death.mod.data$lag14.log.confirmed)#,
#                        # fixed=c(0,0,NA,0,0,0,NA,NA))
# 
# # Now check the residuals
# plot(window(rstandard(death.autoreg)), 
#      ylab = 'Standardized Residuals', type = 'o')
# abline(h=0)
# # ACF of Standardized Residuals
# acf(as.vector(rstandard(death.autoreg)), lag.max = 40,ci.type="ma")
# pacf(as.vector(rstandard(death.autoreg)), lag.max = 40)
# # qqplot
# qqnorm(window(rstandard(death.autoreg)))
# qqline(window(rstandard(death.autoreg)))
# Box.test(residuals(death.autoreg),lag=30,type=c("Ljung-Box"))
```



Predict next week's deaths:
```{r }
pred <- predict(death.mod1,
                newdata = death.mod.data[31:37,5:6], se.fit = TRUE)
preds.real <- data.frame(date = seq(as.Date("2020/04/21"), by = "day", length.out = 7),
                         forecast.deaths = exp(pred$fit), lower = exp(pred$fit-2*pred$se.fit),
                         upper=exp(pred$fit+2*pred$se.fit))
                                                              
deaths <- exp(death.mod.data$log.daily.deaths)

par(mar=c(3,3,3,3)) #change plotting margins to make graph fit
plot(x=death.mod.data$date, y=deaths, 
     type = 'l', col='black',ylab=NA, xlab=NA,
     xlim=as.Date(c("2020-03-15", "2020-04-27")))
lines(preds.real$date, preds.real$forecast.deaths, type= 'l', col='red')
lines(preds.real$date, preds.real$lower, type = 'l', col='blue', lty=5)
lines(preds.real$date, preds.real$upper, type = 'l', col='blue', lty=5)

# par(new = t)
# plot(x=death.mod.data$date,
#      y=exp(death.mod.data$log.daily.confirmed), type = 'l', 
#      col= 'blue',axes = F,ylab=NA)
mtext(side = 1, line = 2, "Date")
mtext(side = 2, line = 2, "Daily Death Count")
mtext(side = 3, line = 2, "Daily Covid 19 Death Forecast")
# mtext(side = 4, line = 2, "Lag 7 Daily Confirmed Case Count")
# axis(side = 4)
legend("topleft",legend=c("Actual Deaths","Predicted Deaths", "+/- 2 Standard Errors"),
       col=c("black","red","blue", "blue"), lty=c(1,1,5,5), inset = 0.03)
```





























